---
title: "Data ETL (CSV Ouput Files)"
author: "BACOU, Melanie"
date: "`r format(Sys.Date(), '%b. %d, %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data ETL (CSV files)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, echo=FALSE}
.pagedtable-wrapper { border: unset; }
.pagedtable-not-empty .pagedtable-footer { border-top: unset; }
.pagedtable-header-name { overflow: unset; }
```

This vignette documents the steps to ingest model output files (in CSV format) for use in the WA+ dashboard. The next [vignette](./03_data-etl-netcdf.html) describes the approach used to ingest and post-process some of the gridded NetCDF output files.

This ETL was validated against two analyses received from IWMI scientists, one for **Mali** (Niger river basin), and another for **Kenya** (Mara river basin). The water accounting models are written in Python and the code repositories are on GitHub (e.g. [WAPORWA notebooks](https://github.com/wateraccounting/WAPORWA) for the Mali analysis).

The results of the WA+ balances consist in time-series of constructed hydrological variables -- the time steps can vary (dekadal, monthly, seasonal, yearly steps). The series are either gridded (NetCDF) or aggregated over an entire zone of interest (CSV output files). The modeled variables are further enriched to construct visual balances (or "accounting sheets") that represent key water availability and water use indicators over time.

The WA+ modeling approach can vary slightly across use cases (different temporal extent, number of sub-basins, different flow units, different mix of input layers and output variables). To the extent possible this ETL intends to produce a generalized data model.

The steps outlined below serve as a blueprint for a generic [data_etl()](../reference/data_etl.html) method included in this package (refer to the method's documentation for more details, esp. how to ingest additional result sets). The objective is to generate a standardized data cube suitable for online analytics (and not limited to this dashboard).


```{r init, echo=FALSE, message=FALSE}
# Load this development package
devtools::load_all(".")

library(rmarkdown)
library(ggplot2)
library(thematic)

# Plot defaults
thematic_on(bg="#f2f1f1", fg=pal[["black"]], accent=pal[["blue"]], 
  font="Roboto Condensed", qualitative=unique(unname(pal)))

th <- function(...) theme(
  legend.position = "none",
  panel.spacing.y = unit(0, "lines"),
  strip.text.y = element_text(size=7, angle=0, hjust=0),
  strip.background = element_blank(),
  axis.text.y = element_blank(),
  axis.title.y = element_blank(),
  axis.ticks.y = element_blank(),
  panel.border = element_blank(),
  panel.grid = element_blank(),
  panel.background = element_blank(),
  ...)
```

```{r}
# Location of WA+ output datasets
dir <- getOption("wa.data")
```

# File and Variable Inventory

We start below by inventorying all basin-level variables contained in the modeled CSV files.

```{r, fig.cap="Kenya - List of output files"}
# List all output CSV
data <- list(
  ken = file.path(dir, "./ken/hydroloop_results/csv"),
  mli = file.path(dir, "/mli/csv_km3")
) %>%
  lapply(list.files, pattern="*.csv", recursive=TRUE, full.names=TRUE) %>%
  lapply(data.table) %>%
  rbindlist(idcol="iso3", use.names=TRUE, fill=TRUE) %>%
  setnames("V1", "path")

# Extract timestamps and sheet codes from file names
data[, `:=`(
  file = basename(path)
)][, `:=`(
  year = str_extract(file, "_[0-9]{4}") %>% str_sub(2,5) %>% as.integer(),
  month = str_extract(file, "[0-9]{4}_[0-9]{1,2}") %>% str_sub(6,7) %>% as.integer(),
  sheet = str_extract(tolower(file), "sheet[0-9]{1}") 
)] %>% setorder(iso3, sheet, year, month, na.last=TRUE)

# Append flow units (typically km3 or MCM, will need to be specified manually)
data[, unit := fcase(
  iso3=="mli", "km3",
  iso3=="ken", "km3"
)]
```

We have 8 yearly water balances for Mali (Sheet #1 variables), 15 yearly balances for Kenya (across `r data[iso3=="ken" & is.na(month), uniqueN(sheet)]` sheets), as well as monthly and seasonal breakdowns. 

```{r, fig.cap="Mali - List of output files"}
data[iso3=="mli", .(iso3, file, sheet, year, month)] %>% 
  paged_table()
```

```{r, fig.cap="Kenya - List of yearly output files"}
data[iso3=="ken" & !is.na(year) & is.na(month), .(iso3, file, sheet, year, month)] %>% 
  paged_table()
```

```{r, fig.cap="Kenya - List of monthly output files"}
data[iso3=="ken" & !is.na(month), .(iso3, file, sheet, year, month)] %>% 
  paged_table()
```

The Kenya analysis also includes yearly, seasonal, and monthly time-series for an additional 57 variables:

```{r, fig.cap="Kenya - List of monthly output files"}
data[iso3=="ken" & is.na(year) & is.na(month), .(iso3, file, sheet, year, month)] %>% 
  paged_table()
```

Let's extract this periodicity (we name it `period`) based on file naming conventions (we ignore Sheets #4 to Sheets #6 for now):

```{r}
data[, `:=`(
  period = fcase(
    is.na(year) & is.na(month) & str_detect(file, "monthly"), "month",
    is.na(year) & is.na(month) & str_detect(file, "season"), "season",
    is.na(year) & is.na(month) & str_detect(file, "year"), "year",
    !is.na(year) & is.na(month), "year",
    !is.na(year) & !is.na(month), "month",
    default=NA
  )
)]

# Verify
data[, .N, keyby=.(iso3, sheet, period)] %>%
  paged_table()
```

Note that in the case of Kenya some of the monthly variables are duplicated (see below).


# Variable Inspection

We also need to validate naming conventions and units for all output variables, so we continue by inspecting individual files starting with Mali.

## Sheet #1

```{r}
f <- data[iso3=="mli"]
mli <- lapply(1:nrow(f), function(x) fread(f[x, path])[, `:=`(
  iso3 = f[x, iso3],
  sheet = f[x, sheet],
  year = f[x, year],
  period = f[x, period]
)]) %>% rbindlist()

setnames(mli, tolower(names(mli))) %>%
  paged_table(options=list(max.print=300))
```

Output for Sheet #1 include `r mli[year==min(year), uniqueN(paste(class, subclass, variable))]` unique variables categorized into `class` and `subclass`. The naming is consistent across years.

```{r, fig.asp=.7}
mli %>% 
  ggplot() +
  geom_line(aes(year, value, color=paste(class, subclass))) +
  facet_grid(paste(class, subclass, variable)~., scales="free_y") +
  th() +
  labs(x=NULL, y=NULL) +
  ggtitle("WA+ Sheet 1 Variables - Mali, 2009 to 2016")

```


Similarly we look at Kenya's monthly time-series. First at Sheet #1 variables:

```{r}
f <- data[iso3=="ken" & period=="month" & sheet=="sheet1" & !is.na(year)]
ken_1_m <- lapply(1:nrow(f), function(x) fread(f[x, path])[, `:=`(
  iso3 = f[x, iso3],
  sheet = f[x, sheet],
  year = f[x, year],
  month = f[x, month],
  period = f[x, period]
)]) %>% rbindlist()

setnames(ken_1_m, tolower(names(ken_1_m))) %>% 
  paged_table(options=list(max.print=300))
```

```{r, fig.asp=.7}
ken_1_m %>% 
  ggplot() +
  geom_line(aes(as.Date(paste(year, month, 1), "%Y %m %d"), value, color=subclass)) +
  facet_grid(paste(class, subclass, variable)~., scales="free_y") +
  th() +
  labs(x=NULL, y=NULL) +
  ggtitle("WA+ Sheet 1 Variables - Kenya, 2003 to 2017")
```

Similarly with the yearly time-series:

```{r}
f <- data[iso3=="ken" & period=="year" & sheet=="sheet1" & !is.na(year)]
ken_1_y <- lapply(1:nrow(f), function(x) fread(f[x, path])[, `:=`(
  iso3 = f[x, iso3],
  sheet = f[x, sheet],
  year = f[x, year],
  month = f[x, month],
  period = f[x, period]
)]) %>% rbindlist()

setnames(ken_1_y, tolower(names(ken_1_y))) %>% 
  paged_table(options=list(max.print=300))
```

## Sheet #2

Then we extract Sheet #2 yearly and monthly variables (this is a breakdown of evapotranspiration by land use classes):

```{r}
f <- data[iso3=="ken" & period %in% c("year", "month") & sheet=="sheet2" & !is.na(year)]
ken_2 <- lapply(1:nrow(f), function(x) fread(f[x, path])[, `:=`(
  iso3 = f[x, iso3],
  sheet = f[x, sheet],
  year = f[x, year],
  month = f[x, month],
  period = f[x, period]
)]) %>% rbindlist()

setnames(ken_2, tolower(names(ken_2))) %>% 
  paged_table(options=list(max.print=300))
```

Sheet #2 uses a "wide" data structure, we first normalize it into a "long" format:

```{r}
# Normalize
ken_2 <- melt(ken_2, id.vars=c("land_use", "class", "iso3", "sheet", "year", "month", "period"),
  variable.factor=FALSE)
# Standardize category names
setnames(ken_2, c("land_use", "class"), c("class", "subclass"))
setorder(ken_2, class, subclass, variable, year, month, period)
```

There are `r ken_2[, .N, by=.(class, subclass, variable)][, .N]` unique variables in this sheet. We only plot variables for the first land use class for illustration purposes:

```{r, fig.asp=.9}
ken_2[period=="month" & class %in% ken_2[, unique(class)][1]] %>% 
  ggplot() +
  geom_line(aes(as.Date(paste(year, month, 1), "%Y %m %d"), value, color=subclass)) +
  facet_grid(paste(class, subclass, variable)~., scales="free_y") +
  th() +
  labs(x=NULL, y=NULL) +
  ggtitle("WA+ Sheet 2 Variables - Kenya, 2003 to 2017")
```


## Sheet #3

We inspect results for Sheet #3 "Agriculture Services". In the Kenya analysis, variables for Sheet #3 to Sheet #6 are available as yearly, seasonal, and/or monthly time-series (across multiple CSV files). The yearly and seasonal files include 1 variable per file. The monthly files include several levels (1.0 to 72.0)[^fn2].

[^fn2]: Check with the authors.

We (arbitrarily) gather all the **seasonal** time-series[^fn1] below:

[^fn1]: What is the difference between variables tagged with "35.0" and "54.0" (e.g. `sheet3_35.0_et_depth_season.csv` and `sheet3_54.0_et_depth_season.csv`)?

```{r}
f <- data[iso3=="ken" & sheet=="sheet3" & period %in% c("season") & str_detect(file, "54.0")]
ken_3 <- lapply(1:nrow(f), function(x) fread(f[x, path])[, `:=`(
  iso3 = f[x, iso3],
  sheet = f[x, sheet],
  year = f[x, year],
  month = f[x, month],
  period = f[x, period],
  variable = str_replace(f[x, file], "sheet3_54.0_", "") %>% str_replace("_season.csv", "")
)]) %>% rbindlist()

ken_3[, `:=`(
  # Drop counter column
  V1 = NULL,
  # Ensure end of season is last day of month (for consistency with other sheets)
  end_dates = end_dates - 1
)][, `:=`(
  # Encode year/month columns using end of season
  year = year(end_dates),
  month = month(end_dates)
)]

setnames(ken_3, 
  c("Seasonal", "start_dates", "end_dates"), c("value", "date_start", "date_end"))
setcolorder(ken_3, 
  c("iso3", "sheet", "variable", "date_start", "date_end", "year", "month", "value"))
setorder(ken_3, variable, year, month) %>% 
  paged_table(options=list(max.print=300))

```

Plotting the entire series (6 variables):

```{r, fig.asp=.5}
ken_3 %>% 
  ggplot() +
  geom_line(aes(as.Date(paste(year, month, 1), "%Y %m %d"), value, color=variable)) +
  facet_grid(variable~., scales="free_y") +
  th() +
  labs(x=NULL, y=NULL) +
  ggtitle("WA+ Sheet 3 Variables - Kenya, 2003 to 2017 (seasonal)")
```


Finally we gather all Sheet #1 to Sheet #3 variables above into a long table format (data cube).

```{r}

cube <- rbind(mli, ken_1_m, ken_1_y, ken_2, ken_3, fill=TRUE)
setcolorder(cube, 
  c("iso3", "sheet", "class", "subclass", "variable", 
    "year", "month", "period", "date_start", "date_end"))

```

For consistency's sake, we encode `date_start` and `date_end` in all sheets (incl. yearly and monthly variables):

```{r}

cube[is.na(date_start) & is.na(month), `:=`(
  # yearly series: 1/1 to 12/31
  date_start = as.IDate(paste(year, 1, 1), "%Y %m %d"),
  date_end = as.IDate(paste(year, 12, 31), "%Y %m %d")
)][is.na(date_start) & !is.na(month), `:=`(
  # monthly series: 1/1 to 1/31
  date_start = as.IDate(paste(year, month, 1), "%Y %m %d"),
  date_end = ceiling_date(as.IDate(paste(year, month, 20), "%Y %m %d"), "months") - days(1)
)]

#saveRDS(cube, "./data-raw/rds/data.rds")

```

At this point the generated data cube includes the following dimensions and measures:

`r paste("*", names(cube), collapse="\n")`


# Calculated Measures

Another ETL step is required to map the variable codes used in the data files to their visual representation in the WA+ Sheets (SVG designs).

```{r, out.width="33%", fig.width=3, fig.show="hold"}
knitr::include_graphics(file.path("./fig", list.files("./fig")))
```

This mapping is embedded in the WA+ `print_sheet` Python module, (e.g. [Sheet 1](https://github.com/wateraccounting/WAPORWA/blob/master/modules/WA/sheet1.py) in the WAPORWA repo). Importantly many SVG elements are **fields calculated** from output variables. We might need to replicate that logic as part of the ETL process[^fn3].

[^fn3]: This is not an ideal approach, we'd rather avoid replicating any hydrological model logic in the visualization layer, and instead export all modeled variables. TBD for a future version.

## Sheet #1

Field mappings and field formulas for Sheet #1 are constructed per the list shown below.

```{r}
s1 <- fread(system.file("./data-raw/csv/sheet_1_schema.csv", package="WADashboard"))
s1[, .SD, .SDcols=-c(1,7)] %>% paged_table()
```

Apply mapping:

```{r}
# Check for NA in data cube
cube[is.na(value), .N, by=.(iso3, sheet, year, month, variable)] %>% paged_table()
```

There are 24 variables with missing values in Sheet #3, we leave them as-is for now.

```{r}
# Add unique variable ID to data cube based on S1 mapping
cube[s1, on=.(sheet, class, subclass, variable), id := i.id]

# Verify
cube[sheet=="sheet1", .N, keyby=.(class, subclass, variable, id)] %>%
  paged_table()

# To wide format
cube_1 <- dcast(cube[sheet=="sheet1"], sheet+iso3+period+date_start+date_end+year+month~id)

# Calculate fields
ids = s1[order(step)][(calculated), unique(id)]
for(i in ids) cube_1[, (i) := s1[id==i, eval(parse(text=formula))]]

# Verify that all Sheet #1 fields have been mapped
s1[!id %in% names(cube_1), id]

# Export
# saveRDS(cube_1, "../data-raw/rds/data_sheet_1.rds")
```

Note that placeholder `Q SW outlet` on Sheet #1 and `land_et` variable in the code need clarification.

## Sheet #2

We proceed similarly for mapping fields in Sheet #2 and Sheet #3. Sheet #2 is a breakdown of ET and T into 280 use classes. Note that we don't have exact formulas to calculate aggregated measures on this sheet (but can be inferred from the SVG design).

```{r}
cube_2 <- cube[sheet=="sheet2"][, id := NULL]
# Wide format
cube_2 <- dcast(cube_2, ...~variable)[, `:=`(
  # Add constructed measures
  evaporation = water + soil + interception,
  et = water + soil + interception + transpiration,
  beneficial = agriculture + environment + economy + energy + leisure
)]
# Long format
cube_2 <- melt(cube_2, id.vars=1:9, variable.factor=FALSE)

# We create unique IDs (needed to populate SVG design)
s2 <- cube_2[, .N, by=.(class, subclass, variable)
][, `:=`(
  N = NULL,
  sheet = "sheet2",
  id = sprintf("%s_%s_%s",
    fcase(
      class=="MANAGED CONVENTIONAL", "malu",
      class=="MANAGED NON_CONVENTIONAL", "malu",
      class=="MODIFIED", "molu",
      class=="PROTECTED", "plu",
      class=="UTILIZED", "ulu"
    ),
    fcase(
      subclass=="Industry", "industry",
      subclass=="Irrigated crops", "crops",
      subclass=="Managed water bodies", "waterbodies",
      subclass=="Others", "others",
      subclass=="Residential", "residential",
      subclass=="Greenhouses", "greenhouses",
      subclass=="Indoor domestic", "idomestic",
      subclass=="Indoor industry", "iiindustry",
      subclass=="Livestock and husbandry", "livestock",
      subclass=="Power and energy", "powerandenergy",
      subclass=="Forest plantations", "forest",
      subclass=="Rainfed crops", "rainfed",
      subclass=="Settlements", "settlements",
      subclass=="Forest", "forest",
      subclass=="Glaciers", "glaciers",
      subclass=="Natural grasslands", "grasslands",
      subclass=="Natural water bodies", "waterbodies",
      subclass=="Shrubland", "shrubland",
      subclass=="Wetlands", "wetlands"
    ),
    fcase(
      variable=="agriculture", "agr",
      variable=="economy", "eco",
      variable=="energy", "ene",
      variable=="environment", "env",
      variable=="interception", "i",
      variable=="leisure", "lei",
      variable=="non_beneficial", "n",
      variable=="soil", "s",
      variable=="transpiration", "t",
      variable=="water", "w",
      variable=="beneficial", "b",
      variable=="et", "et",
      variable=="evaporation", "e"
    ))
)][, id := str_replace(id, "others", fcase(
  # Fix duplicated "others" as in SVG
  class=="MANAGED CONVENTIONAL", "others1",
  class=="MANAGED NON_CONVENTIONAL", "others2",
  default="others")
)]

# Verify ID uniqueness
s2[, .N, by=id][N>1]
```

Apply mapping:

```{r}
# Add unique variable ID to data cube based on S2 mapping
cube_2[s2, on=.(sheet, class, subclass, variable), id := i.id]

# Verify that all Sheet #2 fields have been mapped
s2[!id %in% cube_2[, unique(id)], id]

# Aggregate measure by flow type
vars_1 <- cube_2[, .(value = sum(value, ba.rm=T)
), by=.(sheet, iso3, year, month, period, date_start, date_end, variable)]

# Aggregate measure by top 4 land use class
vars_2 <- cube_2[variable %in% c("et", "transpiration"), .(value = sum(value, na.rm=T)
), by=.(sheet, iso3, year, month, period, date_start, date_end, class, variable)]

# Total ET
vars_3 <- cube_2[variable %in% c("et"), .(value = sum(value, ba.rm=T)
),by=.(sheet, iso3, year, month, period, date_start, date_end, variable)]

# Make unique IDs
vars_1 <- dcast(vars_1, sheet+iso3+period+date_start+date_end+year+month~variable)
vars_3 <- dcast(vars_3, sheet+iso3+period+date_start+date_end+year+month~variable)
vars_2[, `:=`(
  id = sprintf("%s_%s",
    fcase(
      class=="MANAGED CONVENTIONAL", "malu",
      class=="MANAGED NON_CONVENTIONAL", "malu",
      class=="MODIFIED", "molu",
      class=="PROTECTED", "plu",
      class=="UTILIZED", "ulu"
    ),
    fcase(
      variable=="et", "et",
      variable=="transpiration", "t"
    ))
)]
vars_2 <- dcast(vars_2, sheet+iso3+period+date_start+date_end+year+month~id, 
  fun.aggregate=sum, na.rm=T)

# Combine all into wide format
cube_2 <- dcast(cube_2, sheet+iso3+period+date_start+date_end+year+month~id
)[vars_1, on=.(sheet, iso3, period, date_start, date_end, year, month)
][vars_2, on=.(sheet, iso3, period, date_start, date_end, year, month)
][vars_3, on=.(sheet, iso3, period, date_start, date_end, year, month)]

# Verify variable uniqueness
anyDuplicated(names(cube_2))

# Save final Sheet 2 schema
setcolorder(s2, c("sheet", "id", "class", "subclass", "variable"))
# fwrite(s2, "../data-raw/csv/sheet_2_schema.csv")

# Export
# saveRDS(cube_2, "../data-raw/rds/data_sheet_2.rds")
```

Below is a final variable catalog for Sheet #2.

```{r}
s2 <- fread(system.file("./data-raw/csv/sheet_2_schema.csv", package="WADashboard"))
s2[, .SD, .SDcols=-c(1)] %>% paged_table()
```

And we plot some of the aggregate time-series for Kenya.

```{r}

```


## Sheet #3

[forthcoming]


```{r}
cube[sheet=="sheet3", .N, keyby=.(class, subclass, variable)] %>%
  paged_table()
```



In the next [vignette](./03_data-etl-netcdf.html) we outline an approach to ingest all gridded NetCDF output files.

